---
title: "Relationship between ribosomal protein position and tau association in Alzheimer's disease"
author: "Shon Koren"
date: "August 10th, 2020"
output:
  github_document:
        pandoc_args: --webtex
indent: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is code demonstrating the relationship between human ribosomal protein (RP) structural position and association with the microtubule associated protein tau. RP structurual position is based on an "Interface Index" algorithm based on the number of amino acids (AA) interfacing with either rRNA or water (Natchiar et al., 2017, Shigeoka et al., 2019).
&nbsp;
<center>
$$InterfaceIndex = \displaystyle \frac{{\sum AA}_{rRNA}}{\sum(AA_{rRNA}+AA_{water})}$$
</center>

Internal RP proteins have more AA interfacing with rRNA than water resulting in a higher Interface Index. A common cutoff for RP structural position is 0.60 (Shigeoka et al., 2019). Plotting for the relative frequency of RP interface indices (Fig. 1) reveals most RPs are external rather than internal, as expected.

```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE, fig.align = 'center'}
#Initializing data and packages
library(ggplot2)
library(dplyr)
library(ggstatsplot)
setwd("D:/Dropbox (UFL)/Projects/tRNA/Codon Usage")
options(scipen = 999) #Turn off sci-notation

rpf = 'RP.full.csv'
rp.full = as.data.frame(read.csv(rpf, as.is=TRUE,header=TRUE,sep=',', check.names=FALSE, fileEncoding="UTF-8-BOM"))
colnames(rp.full)[1] = 'Symbol'

ggplot(data = rp.full) + geom_histogram(aes(x = Interface, y = stat(count) / sum(count)), binwidth = .05) + xlab("Interface Index of RPs") + ggtitle("Figure 1: Relative frequency of human RP Interface Index") +
  ylab("Relative Frequency") + geom_vline(xintercept = 0.60, color = "red", linetype = "dashed", size = 2) + scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, .15))
```

Next, load the data from Hsieh et al., 2019 which used label free quantification mass spectrometry (LFQ MS) to investigate the tau interactome when immunoprecipitated from human postmortem brain tissue. In this study, tissue samples from the brain of patients with Alzheimer's disease (AD) were compared to control, non-demented patients (n = 4 per group). Before data loading, I filtered out interface index data from any RPs not detected by MS, limiting data to roughly 50 out of 80 potential RPs.

```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE}
h19f = "Hsieh2019.csv"
h19 = as.data.frame(read.csv(h19f, as.is=TRUE,header=TRUE,sep=',', check.names=FALSE, fileEncoding="UTF-8-BOM"))
colnames(h19)[1] = 'Symbol'
h19[is.na(h19)] <- "Other"
h19$Translation[h19$Translation == "TRUE"] = "Translational / Ribosomal"
```

```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE}
ff = 'RP.interface.csv'
rp = as.data.frame(read.csv(ff, as.is=TRUE,header=TRUE,sep=',', check.names=FALSE))
colnames(rp)[1] = 'Symbol'
colnames(rp)[3] = 'AD.CT.FC'
```
As noted by the Hsieh et al., 2019 study among many others including from the Abisambra lab, tau basally associates with RPs. When we plot the data from Hsieh 2019, we see that translational or ribosomal proteins are highly enriched (log2 FC > 2) in AD over control samples (Fig. 2). This suggests that tau has enhanced association with RPs in AD, as noted by other studies.

```{r echo=FALSE, warning = FALSE, error = FALSE, fig.align = 'center'}
ggstatsplot::ggbetweenstats(
  data = h19,
  x = Translation,
  y = AD.CT.FC,
  notch = TRUE,
  mean.ci = FALSE,
  outlier.tagging = FALSE,
  xlab = "Protein Class",
  ylab = "Tau-IP Intensity AD-CT (log2 FC)",
  title = "Figure 2: Tau Association with Translational Proteins in human AD brain",
  bf.message = FALSE,
  results.subtitle = FALSE,
) + geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1)
```

There are two main possibilities dictating tau-RP association ignoring the association of tau to rRNA or other RNA:
1. Tau binds to extra-ribosomal proteins more than RPs in stable ribosomes, or
2. Tau binds to RPs incorporated in ribosomes

If (1) is true, then tau would likely not favor any internal or external RP. If (2) is true, tau would likely favor associating with external RPs and prevent association with more internal RPs.

To assess whether there is a tendency for tau to bind to RPs based on their structural position, the two sample Kolmogorov-Smirnov (two-sample K-S) test can be used to determine whether two distributions are independent of one another. Since a common cutoff for internal RPs is an Interface Index of 0.60, we can separate RPs based on that statistic. The two-sample K-S test can then be used to test whether the tau association of RPs is independent of their structural position in the ribosome:
```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE}

ff = 'RP.interface.csv'
rp = as.data.frame(read.csv(ff, as.is=TRUE,header=TRUE,sep=',', check.names=FALSE, fileEncoding = "UTF-8-BOM"))
colnames(rp)[4] = 'AD.CT.FC'

rp.filt.in = rp %>% filter(Interface > .6)
rp.filt.ex = rp %>% filter(Interface < .6)
ks.test(rp.filt.in$CTmean, rp.filt.ex$CTmean)
```

Since the p value of the two-sample K-S test > 0.05, the two distributions of ribosomal structural position (external, II < 0.60 and internal, II > 0.60) are not independent, suggesting tau associates with RPs regardless of their position in ribosomes. This is more easily interpreted when plotted in (Fig. 3):

```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE, fig.align = 'center'}
p2 = ggstatsplot::ggscatterstats(
  data = rp,
  x = Interface,
  y = CTmean,
  type = "np",
  xlab = "RP Interface Index\nHigher = Interior",
  ylab = "Tau-IP Intensity in Control brain",
  label.var = Symbol,
  label.expression = pval > 10,
  results.subtitle = FALSE,
  marginal = FALSE,
  conf.level = 0,
  title = "Figure 3: Relationship between RP Interface Index and tau association",
  caption = "RP interface data from Natchiar et al., 2017, Shigeoka et al., 2019\nTau-IP data from Hsieh et al., 2019",
  bf.message = FALSE,
  centrality.parameter = "none"
)
p2 + geom_vline(xintercept = 0.60, color = "red", linetype = "dashed", size = 1)

```

This analysis reveals that tau associates with RPs regardless of their structural position. This suggests that (1) is true, tau associates with RPs prior to their incorporation into the full ribosome. However, other factors such as RP-rRNA binding or tau's role in nucleolar activity, rDNA transcription, and ribosome biogenesis can also factor into tau-RP binding. 

Considering that tau associates with RPs more in AD, we can investigate whether tau associates with RPs more in AD as a product of their structural position given what is known about tau's mislocalization and ribosome dysfunction in disease. We can repeat our statistical analysis comparing the distribution of external (<0.60 II) or internal (>0.60 II) RPs associating with tau in AD over in control brain samples (Fig. 4):
```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE,  fig.align = 'center'}
ks.test(rp.filt.in$CTmean, rp.filt.ex$CTmean)

p1 = ggstatsplot::ggscatterstats(
  data = rp,
  x = Interface,
  y = AD.CT.FC,
  type = "np",
  xlab = "RP Interface Index\nHigher = Interior",
  ylab = "Tau-IP Intensity AD-CT (log2 FC)",
  label.var = Symbol,
  label.expression = pval > 1.3,
  results.subtitle = FALSE,
  marginal = FALSE,
  conf.level = 0,
  title = "Figure 4: Relationship between RP interface index and tau association in AD",
  subtitle = "RPs with significant (FDR-pval < 0.05) association with tau in AD over controls have symbols annotated",
  caption = "RP interface data from Natchiar et al., 2017, Shigeoka et al., 2019\nTau-IP data from Hsieh et al., 2019",
  bf.message = FALSE,
  centrality.parameter = "none"
)
p1 + geom_vline(xintercept = 0.60, color = "red", linetype = "dashed", size = 1)

```
However, in doing so we reveal that the enhanced association of tau with RPs also does not correlate with structural position in the ribosome. This supports tau binding to RPs while not incorporated into the full ribosomes.


References:

1.Hsieh, Y.-C. et al. Tau-Mediated Disruption of the Spliceosome Triggers Cryptic RNA Splicing and Neurodegeneration in Alzheimer’s Disease. Cell Rep 29, 301-316.e10 (2019).

2.Natchiar, S. K., Myasnikov, A. G., Kratzat, H., Hazemann, I. & Klaholz, B. P. Visualization of chemical modifications in the human 80S ribosome structure. Nature 551, 472–477 (2017).

3.Shigeoka, T. et al. On-Site Ribosome Remodeling by Locally Synthesized Ribosomal Proteins in Axons. Cell Rep 29, 3605-3619.e10 (2019).
