---
title: "Relationship between ribosomal protein position and tau association in Alzheimer's disease"
author: "Shon A. Koren"
date: "August 10th, 2020"
output:
  github_document:
        pandoc_args: --webtex
indent: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This is code demonstrates the relationship between human ribosomal protein (RP) structural position within the ribosome and association with the microtubule associated protein (MAP) tau. Despite its canonical role as a MAP, recent studies have illustrated the role of tau in (dys)regulating RNA metabolism and protein synthesis (Koren et al., 2020). To assess properties of RPs associated with tau, we can use the "Interface Index" algorithm based on the number of amino acids (AA) interfacing with either rRNA or water for each RP (Natchiar et al., 2017, Shigeoka et al., 2019). 

$$InterfaceIndex = \displaystyle \frac{{\sum AA}_{rRNA}}{\sum(AA_{rRNA}+AA_{water})}$$
We can therefore couple this index with quantitiative measurements of RP-tau association from mass spectrometry (MS) studies of the tau interactome in human brain.


## Most RPs are external, but a defined core of internal RPs exist.
Internal RP proteins have more AA interfacing with rRNA than water resulting in a higher Interface Index. A common cutoff for RP structural position is 0.60 (Shigeoka et al., 2019). Plotting for the relative frequency of RP interface indices (**Fig. 1**) reveals most RPs are external rather than internal, but roughly 30% of RPs reside internally.

```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE, fig.align = 'center'}
#Initializing data and packages
library(ggplot2)
library(dplyr)
library(ggstatsplot)
setwd("D:/Dropbox (UFL)/Projects/tRNA/Codon Usage")
options(scipen = 999) #Turn off sci-notation

rpf = 'RP.full.csv'
rp.full = as.data.frame(read.csv(rpf, as.is=TRUE,header=TRUE,sep=',', check.names=FALSE, fileEncoding="UTF-8-BOM"))
colnames(rp.full)[1] = 'Symbol'

ggplot(data = rp.full) + geom_histogram(aes(x = Interface, y = stat(count) / sum(count)), binwidth = .05) + xlab("Interface Index of RPs") + ggtitle("Figure 1: Relative frequency of human RP Interface Index") +
  ylab("Relative Frequency (%)") + geom_vline(xintercept = 0.60, color = "red", linetype = "dashed", size = 2) + scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, .15))
```

## Tau interactome data from human brain shows high RP association

The most robust investigation of the tau interactome from human brain in AD was recently published (Hsieh et al., 2019). This study used label free quantification mass spectrometry (LFQ MS) to investigate the tau interactome when immunoprecipitated from human postmortem brain tissue. Similar to other earlier studies, Hsieh et al noted that tau associates with many RP and translational proteins and this association is enhanced in AD (**Fig. 2**). However, not all RPs (of roughly 80) were found due to limitations in MS depth, limiting analysis to roughly half (41) of all RPs.

```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE}
h19f = "Hsieh2019.csv"
h19 = as.data.frame(read.csv(h19f, as.is=TRUE,header=TRUE,sep=',', check.names=FALSE, fileEncoding="UTF-8-BOM"))
colnames(h19)[1] = 'Symbol'
h19[is.na(h19)] <- "Other"
h19$Translation[h19$Translation == "TRUE"] = "Translational / Ribosomal"

ggstatsplot::ggbetweenstats(
  data = h19,
  x = Translation,
  y = AD.CT.FC,
  notch = TRUE,
  mean.ci = FALSE,
  outlier.tagging = FALSE,
  xlab = "Protein Class",
  ylab = "Tau-IP Intensity AD-CT (log2 FC)",
  title = "Figure 2: Tau Association with Translational Proteins in human AD brain",
  bf.message = FALSE,
  results.subtitle = FALSE,
) + geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1)
```

## RP structural position and association with tau in control, non-demented brain samples are not related

There are two main possibilities dictating tau-RP association ignoring the association of tau to rRNA or other RNA:  

1. Tau binds to extra-ribosomal proteins more than RPs in stable ribosomes, or  
2. Tau binds to RPs incorporated within ribosomes  

If **(1)** is true, then tau would likely not favor any internal or external RP. If **(2)** is true, tau would likely favor association with external RPs over more internal RPs.

To assess whether there is a tendency for tau to bind to RPs based on their structural position, the two sample Kolmogorov-Smirnov (two-sample K-S) test can be used to determine whether two distributions are independent of one another. Since a common cutoff for internal RPs is an Interface Index of 0.60, we can separate RPs based on that statistic. The two-sample K-S test can then be used to test whether the tau association of RPs is independent of their structural position in the ribosome:
```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE}

ff = 'RP.interface.csv'
rp = as.data.frame(read.csv(ff, as.is=TRUE,header=TRUE,sep=',', check.names=FALSE, fileEncoding = "UTF-8-BOM"))
colnames(rp)[4] = 'AD.CT.FC'

rp.filt.in = rp %>% filter(Interface > .6)
rp.filt.ex = rp %>% filter(Interface < .6)
ks.test(rp.filt.in$CTmean, rp.filt.ex$CTmean)
```

Since the *p* value of the two-sample K-S test > 0.05, the two distributions of ribosomal structural position (external, $II < 0.60$ and internal, $II > 0.60$) are not independent, suggesting tau associates with RPs regardless of their position in ribosomes. This is more easily interpreted when plotted in (**Fig. 3**):

```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE, fig.align = 'center'}
p2 = ggstatsplot::ggscatterstats(
  data = rp,
  x = Interface,
  y = CTmean,
  type = "np",
  xlab = "RP Interface Index\nHigher = Interior",
  ylab = "Tau-IP Intensity in Control brain",
  label.var = Symbol,
  label.expression = pval > 10,
  results.subtitle = FALSE,
  marginal = FALSE,
  conf.level = 0,
  title = "Figure 3: Relationship between RP Interface Index and tau association",
  caption = "RP interface data from Natchiar et al., 2017, Shigeoka et al., 2019\nTau-IP data from Hsieh et al., 2019",
  bf.message = FALSE,
  centrality.parameter = "none"
)
p2 + geom_vline(xintercept = 0.60, color = "red", linetype = "dashed", size = 1)

```

As tau associates with RPs regardless of their structural position, **(2)** is likely false, tau does not associate more with external RPs over internal ones. Instead, tau associates with RPs seemingly regardless of their structural position within the ribosome.

## RP structural position and tau association in AD is also unrelated, despite significantly higher tau-RP association

Considering that tau associates with RPs more in AD, we can investigate whether tau associates with RPs more in AD as a product of their structural position given what is known about tau's mislocalization and ribosome dysfunction in disease. We can repeat our statistical analysis comparing the distribution of external ($II <0.60$) or internal ($II >0.60$) RPs associating with tau in AD over in control brain samples (**Fig. 4**):
```{r echo=FALSE, warning = FALSE, error = FALSE, message=FALSE,  fig.align = 'center'}
ks.test(rp.filt.in$AD.CT.FC, rp.filt.ex$AD.CT.FC)

p1 = ggstatsplot::ggscatterstats(
  data = rp,
  x = Interface,
  y = AD.CT.FC,
  type = "np",
  xlab = "RP Interface Index\nHigher = Interior",
  ylab = "Tau-IP Intensity AD-CT (log2 FC)",
  label.var = Symbol,
  label.expression = pval > 1.3,
  results.subtitle = FALSE,
  marginal = FALSE,
  conf.level = 0,
  title = "Figure 4: Relationship between RP interface index and tau association in AD",
  subtitle = "RPs with significant (FDR-pval < 0.05) association with tau in AD over controls have symbols annotated",
  caption = "RP interface data from Natchiar et al., 2017, Shigeoka et al., 2019\nTau-IP data from Hsieh et al., 2019",
  bf.message = FALSE,
  centrality.parameter = "none"
)
p1 + geom_vline(xintercept = 0.60, color = "red", linetype = "dashed", size = 1)

```

However, similar to in control samples, the increase in tau-RP association in AD does not depend on structural position in the ribosome.\

## References:

1. Hsieh, Y.-C. et al. Tau-Mediated Disruption of the Spliceosome Triggers Cryptic RNA Splicing and Neurodegeneration in Alzheimer’s Disease. Cell Rep 29, 301-316.e10 (2019).\
2. Koren, S. A., Galvis-Escobar, S. & Abisambra, J. F. Tau-mediated dysregulation of RNA: Evidence for a common molecular mechanism of toxicity in frontotemporal dementia and other tauopathies. Neurobiol. Dis. 141, 104939 (2020).\
3. Natchiar, S. K., Myasnikov, A. G., Kratzat, H., Hazemann, I. & Klaholz, B. P. Visualization of chemical modifications in the human 80S ribosome structure. Nature 551, 472–477 (2017).\
4. Shigeoka, T. et al. On-Site Ribosome Remodeling by Locally Synthesized Ribosomal Proteins in Axons. Cell Rep 29, 3605-3619.e10 (2019).


